The extended light profiles of the Sculptor and Ursa Minor dwarf galaxies: Tides or nature?

### 2025-09-02 thesis/abstract.tex
julio sep 1 edits to §1, 2

The satellite galaxies of the Milky Way (MW), with the exception of the
Magellanic clouds, are all dwarf spheroidals (dSphs)---gas-free,
non-rotating stellar systems whos gravity is dominated by dark matter.
The light profiles of dSphs are usually well-approximated by an
exponential law, with a sharp outer cutoff. Yet systems like the
Sculptor (Scl) and Ursa Minor (UMi) dSphs have been found to host member
stars far from the main body, out to \(\sim10\) effective radii. The
origin of these extended profiles is not clear, and may reflect either
the influence of Galactic tides or instead an innate feature. In this
thesis, I review the excestence of an extended light profiles in Scl and
UMi, validating the Bayesian membership catalogue of
\citet{jensen+2024}. To evaluate if tides indeed can explain these
density features, I conduct idealized N-body simulation of both galaxies
in the tidal field of the Milky Way. I find that neigher dwarf has been
subjected to tidal forces strong enough to create the observed extended
light profiles from initially exponential profiles. One complication is
that Sculptor's orbit is strongly influenced by the presence of the
Large Magellanic Cloud (LMC). Because the LMC's mass is poorly
constrained, Sculptor's long term orbital history is highly uncertain.
Despite this uncertainty, our models still suggest that the combined
tides of the LMC and MW are still unable to explain Scl's outer profile
over the past 5 Gyr. I conclude that the extended light profiles of Scl
and UMi are innate, likely reflecting their formation histories.
Mergers, accretion events, or complex, multi-component star formation
episodes are possible explanations for such an extended outer profile.

>>>>

The satellite galaxies of the Milky Way (MW), with the exception of the
Magellanic Clouds, are all dwarf spheroidals (dSphs)---gas-free,
non-rotating stellar systems whose gravity is dominated by dark matter.
The light profiles of dSphs are usually well-approximated by an
exponential law, with a sharp outer cutoff. Yet systems like the
Sculptor (Scl) and Ursa Minor (UMi) dSphs have been found to host member
stars far from the main body, out to \(\sim10\) effective radii,
indicative of outer deviations from an exponential law. The origin of
these extended profiles is not clear, and may reflect either the
influence of Galactic tides or, instead, an innate feature. In this
thesis, we review the existence of extended light profiles in Scl and
UMi, validating the Bayesian membership catalogue of
\citet{jensen+2024}. To evaluate if tides indeed can explain these
density features, we conduct idealized N-body simulation of both
galaxies in the tidal field of the Milky Way. We find that neither dwarf
has been subjected to tidal forces strong enough to create the observed
extended light profiles from initially exponential profiles. One
complication is that Sculptor's orbit is strongly influenced by the
presence of the Large Magellanic Cloud (LMC). Because the LMC's mass is
poorly constrained, Sculptor's long term orbital history is highly
uncertain. Despite this uncertainty, our models still suggest that the
combined tides of the LMC and MW are unable to explain Scl's outer
profile. We conclude that the extended light profiles of Scl and UMi are
innate, likely reflecting their assembly histories. Mergers, accretion
events, or complex, multi-component star formation episodes are possible
explanations for the observations ed extended light profiles.

### 2025-09-02 thesis/data.tex
julio sep 1 edits to §1, 2

As discussed in Section~\ref{sec:exponential_profiles}, the projected
luminosity/stellar mass density profile of dwarf galaxies is generally
well-described by an exponential law (Eq.~\ref{eq:exponential_law}). A
prototypical example is the Fornax dSph --- well-fit by an exponential
over 4.5 decades in surface density. On the other hand, dSph like
Sculptor or Ursa Minor have profiles with deviate significantly from an
exponential profile fitted to the inner regions, with a clear excess of
stars/light in the outer regions. In this chapter, we critically review
the density profile of MW classical dwarfs. This thesis concerns the
origin of this excess, with a tidal interpretation explored in
subsequent chapters.

@@

Measuring the light profile of a resolved galaxy requires careful
consideration of whether a given star belongs to the system. Without
removing contamination from foreground/background features, faint
features may be lost in the noise or be of uncertain association. When
only photometric data was available, the galaxy membership of stars were
ascertained using the colour-magnitude diagram alone \citep[e.g.,
matched filter methods like][]{rockosi+2002}. Now that \emph{Gaia} data
are available, stellar parallax and proper motion are also available to
improve membership assignment.

Here, we use the \citet{jensen+2024}'s (hereafter J+24) membership
probabilities for \emph{Gaia} observations. J+24 used a Bayesian
framework incorporating proper motion (PM), colour-magnitude diagram
(CMD), and spatial information to determine the probability that a given
star belongs to the satellite or foreground/background. By accounting
for PM in particular, J+24 produced low contamination samples of
candidate member stars out to large distances from a dwarf galaxy. J+24
extended the algorithm presented in \citet{MV2020a}; \citet{MV2020b} by
additionally includes a secondary, extended spatial component. J+24
detected candidate members out to \textasciitilde10 half-light radii
from the centres of some galaxies (\(R_h\)). Similar recent work has
also included \citet{pace+li2019}; \citet{battaglia+2022};
\citet{pace+erkal+li2022}; \citet{qi+2022}.

J+24 construct likelihoods, \({\cal L}\), representing the probability
density that a star is consistent with either the foreground/background,
\({\cal L}_{\rm bg}\), or the satellite galaxy, \({\cal L}_{\rm sat}\).
In either case, the likelihoods are the product of a spatial, PM, and
CMD component, \begin{equation}{
{\cal L} = {\cal L}_{\rm space}\ {\cal L}_{\rm PM}\ {\cal L}_{\rm CMD}.
}\end{equation}

For the satellite, the \emph{spatial likelihood} is specified either as
a one- or two-component elliptical exponential profile. A second
component is included only when the preferred amplitude of the second
component is non-zero. The \emph{proper-motion likelihood} quantifies
the agreement of a star's motion with the dwarf galaxy's systemic
motion, accounting for observational uncertainties. The \emph{CMD
likelihood} measures the consistency of a star's \emph{Gaia} \(G\)
magnitude and \(G_{\rm BP}- G_{\rm RP}\) colour with theoretical
isochrones for the galaxy. For the background likelihoods, the spatial
likelihood is a uniform distribution over the field, and the background
CMD and PM likelihoods are constructed empirically from the stars in an
annulus far from the satellite. Each likelihood is normalized as a
probability density over the respective parameter space.

@@

For our fiducial sample, we adopt a minimum probability of
\(P_{\rm sat} = 0.2\). We use the \(P_{\rm sat}\) values from the
elliptical 2-component runs if a galaxy showed evidence for an outer
component, the 1-component run otherwise. Most stars have
\(P_{\rm  sat}\) values which are nearly 0 or 1, so the exact choice of
probability threshold has little effect on the resulting sample. Even at
our relatively generous probability threshold of 0.2, the purity remains
high when validated against spectroscopic line-of-sight (LOS) velocities
(\textasciitilde90\%, J+24). We note, however, that these purity
estimates may be biased. Stars with LOS velocities are typically
brighter and, as a result, have more precise \emph{Gaia} measurements.
However, we find that our conclusions are unchanged when limiting
samples to only the brightest stars. Altogether, the J+24 method
provides a high-quality, low-contamination sample of dwarf galaxy
candidate member stars, which we will now investigate.

@@

We analyze stellar distributions in the tangent plane, considering as
well the projected shape of a galaxy. The tangent plane coordinates
\(\xi\) and \(\eta\) are offsets in RA and declination as measured on
the plane tangent to the galaxy centre. To account for the elliptical
shape of the galaxy, we use \(R_{\rm ell}\), which we define as the
circularized elliptical radius, \begin{equation}{
R_{\rm ell}^2 = a\,b\,\left(\frac{{\xi'}^2}{a^2} + \frac{{\eta'}^2}{b^2} \right),
}\end{equation} where \(\xi'\) and \(\eta'\) are the tangent plane
coordinate rotated to align with a dwarf galaxy's major and minor axis,
and \(a\) and \(b\) are the semi-major and semi-minor axis of the
galaxy.

We illustrate the likelihood-based membership selection criteria through
a progressive tightening of criteria. First, we consider a
minimally-refined sample, \textbf{all}, only excluding stars with poor
astrometry, unreliable photometry, or inconsistent parallaxes. Next,
incorporating CMD and PM information, we use the \textbf{CMD+PM} to test
a selection method agnostic to the spatial position. This sample
includes stars where the CMD and PM combined likelihood favours
satellite membership:
\begin{equation}\protect\phantomsection\label{eq:sel_cmd_pm}{
{\cal L}_{\rm CMD,\ sat}\ {\cal L}_{\rm PM,\ sat} > {\cal L}_{\rm CMD,\ bg}\ {\cal L}_{\rm PM,\ bg}.
}\end{equation} Next, our \textbf{fiducial} sample,
\(P_{\rm sat} > 0.2\), includes a spatial likelihood as described in
J+24. Finally, the \textbf{RV members} (for radial velocity) sample adds
line-of-sight velocity information (see Appendix \ref{sec:rv_obs} for
details). While the latter is the best motivated membership sample, we
do not use this sample for stellar density analysis due to its
incompleteness and complex selection function.

Figs.~\ref{fig:scl_selection}, \ref{fig:umi_selection}, \ref{fig:fornax_selection}
shows our fiducial sample for each galaxy, and the impact of alternate
samples, in tangent-plane coordinates, the CMD, and proper-motion space.
The ``all'' sample extends uniformly across the tangent plane, but
includes a substantial background population. The ``CMD+PM'' sample has
a much lower background density in the tangent plane, revealing the
satellite more clearly. The fiducial sample appears similar to the
``CMD+PM'' sample in the CMD and PM planes, but excludes improbably
distant stars. The ``RV member'' sample also traces out similar
distributions in CMD and PM space as the fiducial sample, with less
dispersion, likely reflecting the brighter magnitudes of stars with
spectroscopic follow-up. Each selection criteria---spatial, CMD, and
PM---contributes towards a high-quality membership criterion.

@@

One limitation of the J+24 method is the assumption of a specific
density profile for the dwarf galaxy (a one- or two-component
exponential), which may impact the membership probability of distant
stars. However, there are no clear extensions or overdensities in the
``CMD+PM'' selected sample, which does not include a spatial likelihood.
If even fainter, more extended features exist, they are apparently not
detectable with \emph{Gaia}.

To illustrate the extent of Sculptor and Ursa Minor, we highlight
distant, spectroscopically confirmed members with red-outlined stars in
Figs.~\ref{fig:scl_selection}, \ref{fig:umi_selection}. In particular,
\citet{sestito+2023a}; \citet{sestito+2023b} targeted distant, bright
stars from the J+24 candidate membership list. The most distant stars
confirmed in their work lie at radii \(\gtrsim 10 R_h\) from the centres
of each dwarf. For an exponential profile, 99.95\% of stars fall within
a radius of \(6R_h\), so the presence of very distant member stars
provides evidence for deviations from an exponential profile.

@@

Fig.~\ref{fig:scl_observed_profiles} show the derived density profiles
for Sculptor, Ursa Minor, and Fornax. We calculate density profiles for
three different samples from above: ``all'', ``CMD+PM'', and the
``fiducial'' sample. In each case, all samples are the same towards the
inner regions of the satellite. However, the density profile from
``all'' plateaus at the total background in the field at radii of 30-60
arcminutes, depending on the galaxy. By restricting stellar membership
with CMD + PM, the background is reduced by 1-2 dex. The ``CMD+PM''
background plateau represents the density of background stars which
could be mistaken as members because of their coincident colours and
proper motions. Finally, the ``all -- background'' profile results from
subtracting the apparent background in the ``all'' profile.

While the density profiles of all samples agrees in the inner regions,
they begin to deviate outside \(6 R_h\). Outside this radius, the
fiducial profile extends 1--2 magnitudes below the CMD+PM selection
background. Because the fiducial profile in this region depends
critically on the spatial likelihood, its shape is vulnerable to the
assumed density profile (an exponential in J+24). We explore this in
more detail in Appendix \ref{sec:density_extra}.

Near to UMi, there is a small (\(R_h\sim 0.5'\)), likely unassociated,
ultrafaint star cluster, Muñoz 1 \citep{munoz+2012}. The cluster is at a
relative position of \((\xi, \eta) \approx(-42, -15)\) arcminutes,
corresponding to an elliptical radius of 37 arcminutes. This cluster
does not have a bright RGB, so has few stars brighter than a magnitude
\(G=22\). The cluster has little effect on the elliptically-averaged
density profile (see location on Fig.~\ref{fig:scl_observed_profiles}).

Fig.~\ref{fig:classical_dwarfs_densities} compares the fiducial density
profiles of Sculptor, Ursa Minor, Fornax, and other classical dwarf
galaxies. Of the classicals, we exclude Antlia II, due to the extremely
high background, and Sagittarius, which was not included in J+24. The
density profiles are scaled to match at the half-light radius from
\citet{munoz+2018}. All of the classical dwarfs appear to be well
described by an exponential profile in the inner regions. In the outer
regions, however, Sculptor and Ursa Minor deviate, and show a clear
outer excess over an exponential law (solid black line). These galaxies
are better fit by a Plummer law (dashed black line). The deviation from
an exponential grows outwards, and at \(\sim 8 R_h\), may reach 2 orders
of magnitude. The remainder of this thesis will be devoted to assessing
whether the outer excess shown by Scl and UMi are due to Galactic tides.

>>>>

As discussed in Section~\ref{sec:exponential_profiles}, the projected
luminosity/stellar mass density profile of dwarf galaxies is generally
well-described by an exponential law (Eq.~\ref{eq:exponential_law}). A
prototypical example is the Fornax dSph --- well-fit by an exponential
over 4.5 decades in surface density. On the other hand, dSphs like
Sculptor or Ursa Minor have profiles which deviate significantly from an
exponential profile fitted to the inner regions, with a clear excess of
stars/light in the outer regions. In this chapter, we critically review
the density profile of the MW classical dwarfs. This thesis concerns the
origin of this excess, with a tidal interpretation explored in
subsequent chapters.

@@

Measuring the light profile of a resolved galaxy requires careful
consideration of whether any given star belongs to the system or not.
Without removing contamination from foreground/background sources, faint
features may be lost in the noise or be of uncertain association. When
only photometric data was available, the membership of stars was
ascertained using the colour-magnitude diagram alone \citep[e.g.,
matched filter methods like those used by][]{rockosi+2002}. Now that
\emph{Gaia} data are available, stellar parallax and proper motion are
also available to improve membership assignment.

Here, we use the \citet{jensen+2024}'s (hereafter J+24) membership
probabilities from \emph{Gaia} data. J+24 used a Bayesian framework
incorporating proper motion (PM), colour-magnitude diagram (CMD), and
spatial information to determine the probability that a given star
belongs to the satellite or foreground/background. By accounting for PM
in particular, J+24 produced low contamination samples of candidate
member stars out to large distances from a dwarf galaxy. J+24 extended
the algorithm presented in \citet{MV2020a}; \citet{MV2020b} by
additionally including a secondary, extended spatial component. J+24
detected candidate members out to \textasciitilde10 half-light radii
from the centres of some galaxies (\(R_h\)). Similar recent work has
also included \citet{pace+li2019}; \citet{battaglia+2022};
\citet{pace+erkal+li2022}; \citet{qi+2022}.

J+24's method uses likelihoods, \({\cal L}\), representing the
probability density that a star is consistent with either the
foreground/background, \({\cal L}_{\rm bg}\), or the satellite galaxy,
\({\cal L}_{\rm sat}\). In either case, the likelihoods are the product
of a spatial, PM, and CMD component, \begin{equation}{
{\cal L} = {\cal L}_{\rm space}\ {\cal L}_{\rm PM}\ {\cal L}_{\rm CMD}.
}\end{equation}

For a satellite, the \emph{spatial likelihood} is specified either as a
one- or two-component elliptical exponential profile. A second component
is included only when the preferred amplitude of the second component is
non-zero. The \emph{proper-motion likelihood} quantifies the agreement
of a star's motion with the dwarf galaxy's systemic motion, accounting
for observational uncertainties. The \emph{CMD likelihood} measures the
consistency of a star's \emph{Gaia} \(G\) magnitude and
\(G_{\rm BP}- G_{\rm RP}\) colour with theoretical isochrones for the
galaxy. For the background likelihoods, the spatial likelihood is a
uniform distribution over the field, and the background CMD and PM
likelihoods are constructed empirically from the stars in an annulus far
from the satellite. Each likelihood is normalized as a probability
density over the respective parameter space.

@@

For our fiducial sample, we adopt a minimum probability of
\(P_{\rm sat} = 0.2\). We do not filter on magnitudes explicitly, but
J+24's quality cuts typically only include stars with \(G < 21\). We use
the \(P_{\rm sat}\) values from the elliptical 2-component runs if a
galaxy shows evidence for an outer component, the 1-component run
otherwise. Most stars have \(P_{\rm  sat}\) values which are nearly 0 or
1, so the exact choice of probability threshold has little effect on the
resulting sample. Even at our relatively generous probability threshold
of 0.2, the purity remains high when validated against spectroscopic
line-of-sight (LOS) velocities (\textasciitilde90\%, J+24). We note,
however, that these purity estimates may be biased. Stars with LOS
velocities are typically brighter and, as a result, have more precise
\emph{Gaia} measurements. However, we find that our conclusions are
unchanged when limiting samples to only the brightest stars. Altogether,
the J+24 method provides a high-quality, low-contamination sample of
dwarf galaxy candidate member stars, which we will now investigate.

@@

We analyze stellar distributions in the tangent plane, considering as
well the projected shape of a galaxy. The tangent plane coordinates
\(\xi\) and \(\eta\) are offsets in RA and declination as measured on
the plane tangent relative to the galaxy centre. To account for the
elliptical shape of the galaxy, we use \(R_{\rm ell}\), which we define
as the circularized elliptical radius, \begin{equation}{
R_{\rm ell}^2 = a\,b\,\left(\frac{{\xi'}^2}{a^2} + \frac{{\eta'}^2}{b^2} \right),
}\end{equation} where \(\xi'\) and \(\eta'\) are the tangent plane
coordinate rotated to align with a dwarf galaxy's major and minor axis,
and \(a\) and \(b\) are the semi-major and semi-minor axis of the
galaxy.

We illustrate the likelihood-based membership selection criteria through
a progressive tightening of criteria. First, we consider a
minimally-refined sample, \textbf{all}, which only excludes stars with
poor astrometry, unreliable photometry, or inconsistent parallaxes.
Next, incorporating CMD and PM information, we use the \textbf{CMD+PM}
to test a selection method agnostic to the spatial position. This sample
includes stars where the CMD and PM combined likelihood favours
satellite membership:
\begin{equation}\protect\phantomsection\label{eq:sel_cmd_pm}{
{\cal L}_{\rm CMD,\ sat}\ {\cal L}_{\rm PM,\ sat} > {\cal L}_{\rm CMD,\ bg}\ {\cal L}_{\rm PM,\ bg}.
}\end{equation} Next, our \textbf{fiducial} sample,
\(P_{\rm sat} > 0.2\), includes a spatial likelihood as described in
J+24. Finally, the \textbf{RV members} (for radial velocity) sample adds
line-of-sight velocity information (see Appendix \ref{sec:rv_obs} for
details). While the latter is the best motivated membership sample, we
do not use this sample for stellar density analysis due to its
incompleteness and complex selection function.

Figs.~\ref{fig:scl_selection}, \ref{fig:umi_selection}, \ref{fig:fornax_selection}
show our fiducial sample for three dSphs (Sculptor, Ursa Minor, and
Fornax), as well as impact of different samples, in tangent-plane
coordinates, the CMD, and proper-motion space. The ``all'' sample
extends uniformly across the tangent plane, but includes a substantial
background population. The ``CMD+PM'' sample has a much lower background
density in the tangent plane, revealing the satellite more clearly. The
fiducial sample appears similar to the ``CMD+PM'' sample in the CMD and
PM planes, but excludes improbably distant stars. The ``RV member''
sample also traces out similar distributions in CMD and PM space as the
fiducial sample, with less dispersion, likely reflecting the brighter
magnitudes of stars with spectroscopic follow-up. Each selection
criteria---spatial, CMD, and PM---contributes towards a high-quality
membership assignment.

@@

One limitation of the J+24 method is the assumption of a specific
density profile for the dwarf galaxy (a one- or two-component
exponential), which may impact the membership probability of distant
stars. However, there are no clear extensions or overdensities in the
``CMD+PM'' selected sample, which does not include a spatial likelihood.
If even fainter, more extended features exist, they are not clearly
detectable with \emph{Gaia}.

To illustrate the extent of Sculptor and Ursa Minor, we highlight
distant, spectroscopically confirmed members with red-outlined stars in
Figs.~\ref{fig:scl_selection}, \ref{fig:umi_selection}. In particular,
\citet{sestito+2023a}; \citet{sestito+2023b} targeted distant, bright
stars from the J+24 candidate membership list. The most distant stars
confirmed in their work lie at radii \(\gtrsim 10 R_h\) from the centres
of each dwarf. For an exponential profile, 99.95\% of stars fall within
a radius of \(6R_h\), so the mere presence of very distant member stars
provides evidence for deviations from an exponential profile.

@@

Fig.~\ref{fig:scl_observed_profiles} show the derived density profiles
for Sculptor, Ursa Minor, and Fornax. We calculate density profiles for
three different samples from above: ``all'', ``CMD+PM'', and the
``fiducial'' sample. In each case, all samples coincide towards the
inner regions of the satellite. However, the density profile from
``all'' plateaus at the total background in the field at radii of 30-60
arcminutes, depending on the galaxy. By restricting stellar membership
with CMD + PM, the background is reduced by 1-2 dex. The ``CMD+PM''
background plateau represents the density of background stars which
could be mistaken as members because of their coincident colours and
proper motions. Finally, the ``all -- background'' profile results from
subtracting the apparent background in the ``all'' profile.

While the density profiles of all samples agree in the inner regions,
they begin to deviate outside \(6 R_h\). Outside this radius, the
fiducial profile extends 1--2 magnitudes below the CMD+PM selection
background. Because the fiducial profile in this region depends
critically on the spatial likelihood, its shape is vulnerable to the
assumed density profile (an exponential in J+24). We explore this in
more detail in Appendix \ref{sec:density_extra}.

Near UMi, there is a small (\(R_h\sim 0.5'\)), likely unassociated,
ultrafaint star cluster, Muñoz 1 \citep{munoz+2012}. The cluster is at a
relative position of \((\xi, \eta) \approx(-42, -15)\) arcminutes,
corresponding to an elliptical radius of 37 arcminutes. This cluster
does not have a bright RGB, so has few stars brighter than a magnitude
\(G=22\). The cluster has little effect on the elliptically-averaged
density profile (see location on Fig.~\ref{fig:scl_observed_profiles}).

Fig.~\ref{fig:classical_dwarfs_densities} compares the fiducial density
profiles of Sculptor, Ursa Minor, Fornax, and other classical dwarf
galaxies. Of the classicals, we exclude Antlia II, due to the extremely
high background, and Sagittarius, which was not included in J+24. The
density profiles are scaled to match at the half-light radius, taken
from \citet{munoz+2018}. All of the classical dwarfs appear to be well
described by an exponential profile in the inner regions. In the outer
regions, however, Sculptor and Ursa Minor deviate, and show a clear
outer excess over an exponential law (solid black line). These galaxies
are better fit by a Plummer law (dashed black line). The deviation from
an exponential grows outwards, and at \(\sim 8 R_h\), may reach 2 orders
of magnitude. The remainder of this thesis will be devoted to assessing
whether the outer excess shown by Scl and UMi are due to Galactic tides.
